---
title: "Activity monitoring data analysis"
author: "Astrid Babayan"
date: "August 21, 2016"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

```

##Loading and preprocessing the data
### 1. Code for reading in the dataset and/or processing the data.


```{r read file, echo=TRUE}

library(mice)
library(VIM)

setwd("U:/R/COurse5/Week2")
df.activity<-read.csv("activity.csv")

summary(df.activity)

```

##What is mean total number of steps taken per day?
### 2. Histogram of the total number of steps taken each day


```{r steps, echo=TRUE}

daily.aggregate<-aggregate(x=df.activity$steps, by=list(date=df.activity$date), FUN=sum)

hist(daily.aggregate$x, main="Daily Steps", xlab="Steps", col="red")

```


### 3. Mean and median number of steps taken each day


Average daily step count is `r round(mean(daily.aggregate$x, na.rm=TRUE), digits=1)`.

Median daily step count is `r median(daily.aggregate$x, na.rm=TRUE)`.

##What is the average daily activity pattern?



``` {r 5-min interval aggregation, echo=TRUE}

interval.aggregate<-aggregate(x=df.activity$steps, by=list(interval=df.activity$interval), FUN="mean", na.rm=TRUE)

hist(interval.aggregate$x,  main="Average Interval Steps", xlab="Average number of steps per interval", col="blue")
```


### 4. Time series plot of the average number of steps taken


```{r time series plot, echo=TRUE}
plot(x=interval.aggregate$interval, y=interval.aggregate$x, type="l", xlab="Interval", ylab="Average Number of Steps", main="Average Number of Steps by Interval", ylim=c(0,250))
with(subset(interval.aggregate, x==max(interval.aggregate$x)),text(interval,x,"*", col="red"))
```


### 5. The 5-minute interval that, on average, contains the maximum number of steps


```{r maximum interval, echo=TRUE}

maximum<-subset(interval.aggregate, x==max(interval.aggregate$x))
       
```


It occures at 5-minute interval number `r maximum$interval`. The average number of steps at that interval is `r round(maximum$x, 1)`. 


##Imputing missing values

### 6. Code to describe and show a strategy for imputing missing data


``` {r missing data pattern, echo=TRUE}
mice_plot <- aggr(df.activity, col=c('navyblue','yellow'),
                  numbers=TRUE, sortVars=TRUE,
                  labels=names(df.activity), cex.axis=.7,
                  gap=3, ylab=c("Missing data","Pattern"))
```

There are `r sum(is.na(df.activity$steps))` missing values in the dataset for variable "steps".

Missing values are imputed by using "mice" library's pmm method (predictive mean method).

```{r imputing missing values for steps, echo=TRUE}

imputed_Data <- mice(data = df.activity, m = 1, method = "pmm", maxit = 10, seed = 100)

imputed_df.activity<-complete(imputed_Data, 1)
```

Missing data pattern plot below confirms the imputation.

``` {r imputed data pattern, echo=TRUE}
mice_plot <- aggr(imputed_df.activity, col=c('navyblue','yellow'),
                  numbers=TRUE, sortVars=TRUE,
                  labels=names(imputed_df.activity), cex.axis=.7,
                  gap=3, ylab=c("Missing data","Pattern"))

```


### 7. Histogram of the total number of steps taken each day after missing values are imputed


```{r imputed steps, echo=TRUE}

imp_daily.aggregate<-aggregate(x=imputed_df.activity$steps, by=list(date=imputed_df.activity$date), FUN=sum)

hist(imp_daily.aggregate$x, main="Daily Steps", xlab="Steps", col="red")

```


Average daily step count after imputation is `r round(mean(imp_daily.aggregate$x, na.rm=TRUE), digits=1)`.

Median daily step count after imputation is `r median(imp_daily.aggregate$x, na.rm=TRUE)`.

The average values did not change much. This is because we used "pmm" method (predictive mean method). Median is affected by imputation. The new median is higher.


##Are there differences in activity patterns between weekdays and weekends?

### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r weekedn factor variable, echo=TRUE}
#creating weekend indicator variable

imputed_df.activity$weekend<-ifelse(weekdays(as.Date(imputed_df.activity$date)) %in% c("Saturday", "Sunday"), "YES", "NO")
```


```{r panel plot by weekday, echo=TRUE, fig.width=8, fig.height=7}
#calculating average step values by intervals

imputed_daily.aggregate<-aggregate(x=imputed_df.activity$steps, 
                                   by=list(interval=imputed_df.activity$interval, weekend=imputed_df.activity$weekend), FUN=mean)


par(mfrow=c(2,1))

plot(subset(imputed_daily.aggregate, weekend=="YES")$interval, subset(imputed_daily.aggregate, weekend=="YES")$x,
     type="l",
     main="Weekend", 
     xlab="Interval",
     ylab="Average Number of Steps",
     ylim=c(0,220),
     col="blue")
plot(subset(imputed_daily.aggregate, weekend=="NO")$interval, subset(imputed_daily.aggregate, weekend=="NO")$x,
     main="Weekdays", 
     type="l",
     xlab="Interval", 
     ylab="Average Number of Steps",
     ylim=c(0,220),
     col="red")

```

Average number of steps on weekeds are lower between 500 and 1000 intervals but slightly higher on the later intervals.


